%% sol-ch04.tex — Solutions for Chapter 4: Nondeterministic Finite Automata

\chapter{Nondeterministic Finite Automata --- Solutions}

\begin{enumerate}[{4.}1.]

%% 4.1
\item Because Figure~4.25 is a printed figure, we describe the subset-construction
process generically.  For each NDFA $A=\langle\Sigma,S,S_0,\delta,F\rangle$
with $\Sigma=\{\pmb{a},\pmb{b},\pmb{c}\}$, we build $A^d$ by:
\begin{enumerate}[(1)]
  \item Taking the start state of $A^d$ to be $S_0$ (treated as a single state
        representing the set of start states).
  \item For each state $Q\in\wp(S)$ of $A^d$ and each $\pmb{\sigma}\in\Sigma$,
        setting $\delta^d(Q,\pmb{\sigma})=\bigcup_{q\in Q}\delta(q,\pmb{\sigma})$.
  \item Declaring $Q$ final iff $Q\cap F\neq\emptyset$.
  \item Retaining only the states reachable from $S_0$.
\end{enumerate}
The resulting DFAs for each part of Figure~4.25 are obtained by applying this
procedure; since the figures are referenced externally, the specific transition
tables depend on the transitions shown in the printed text.

%% 4.2
\item Consider Example~4.17: a $\lambda$-NDFA $E$ identical to the NDFA $D$
of Example~4.13 but with the extra transition $\delta(s_1,\lambda)=\{s_2\}$.

\begin{enumerate}[(a)]
  \item \emph{Remove $\lambda$-transitions (Definition~4.9).}  Apply
  $\delta^{\sigma}_{\lambda}$: for each state $s$ and each input letter
  $\pmb{a}\in\Sigma$, set
  $\delta'(s,\pmb{a})=\DBAR_{\lambda}(s,\pmb{a})=\Lambda(\bigcup_{t\in\Lambda(s)}\delta_\lambda(t,\pmb{a}))$.
  The $\lambda$-closures are:
  \[
    \Lambda(s_0)=\{s_0\},\quad \Lambda(s_1)=\{s_1,s_2\},\quad \Lambda(s_2)=\{s_2\},\quad \Lambda(s_3)=\{s_3\}.
  \]
  (The $\lambda$-transition $s_1\to s_2$ means $s_2\in\Lambda(s_1)$.)
  The resulting NDFA $A^{\sigma}_\lambda$ without $\lambda$-transitions has the
  same state set $\{s_0,s_1,s_2,s_3\}$ (plus the auxiliary start state $q_0$
  if $\lambda$ is not accepted) and transitions derived from the formula above.

  \item \emph{Convert to DFA (Definition~4.5).}  Starting from the state
  $\{s_0\}$ (or the appropriate initial set), compute:
  \[
    \delta^d(\{s_0\},\pmb{a}),\quad \delta^d(\{s_0\},\pmb{b}),\quad\ldots
  \]
  and continue until closure is reached.  Since the exact transitions of $E$
  depend on the printed Figure~4.23, the reader should apply the subset
  construction to the NDFA $A^{\sigma}_\lambda$ produced in part~(a),
  grouping into equivalence classes of reachable state-sets.
\end{enumerate}

%% 4.3
\item Consider the NDFA $A$ from Example~4.4:
$A=\langle\{\pmb{a},\pmb{b}\},\{s,r,t\},\{s\},\delta,\{r,t\}\rangle$
with $\delta(s,\pmb{a})=\{s,r\}$, $\delta(s,\pmb{b})=\{s,t\}$,
$\delta(r,\pmb{a})=\delta(r,\pmb{b})=\delta(t,\pmb{a})=\delta(t,\pmb{b})=\emptyset$.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no \textless SOS\textgreater{} or
    \textless EOS\textgreater).}  Encode states with two flip-flops:
    $s=00$, $r=01$, $t=10$.  The next-state logic computes
    $D_0,D_1$ as functions of the current state bits and the input.
    Because the machine is nondeterministic, the circuit uses OR gates to
    collect contributions from multiple possible transitions.
    (The precise gate-level diagram depends on the chosen encoding and
    hazard-minimization strategy; the reader should apply the standard
    NDFA circuit procedure from Section~4.4 of the text.)

  \item \emph{Circuit diagram (with \textless SOS\textgreater{} and
    \textless EOS\textgreater).}  Add dedicated SOS and EOS inputs that
    respectively force the machine into its start state and gate the
    final-state outputs, following the standard encoding conventions of
    Chapter~4.

  \item \emph{DFA via subset construction (connected portion only).}
  The start state of $A^d$ is $\{s\}$.  Compute:
  \begin{center}\begin{tabular}{c|cc}
    State & $\pmb{a}$ & $\pmb{b}$ \\\hline
    $\{s\}$ & $\{s,r\}$ & $\{s,t\}$ \\
    $\{s,r\}$ & $\{s,r\}$ & $\{s,t\}$ \\
    $\{s,t\}$ & $\{s,r\}$ & $\{s,t\}$ \\
  \end{tabular}\end{center}
  Final states (sets intersecting $\{r,t\}$): $\{s,r\}$ and $\{s,t\}$.
  The dead state $\emptyset$ is not reachable.  The connected DFA has
  three states.

  \item \emph{DFA including disconnected portion.}
  The full set $\wp(\{s,r,t\})=\{\emptyset,\{s\},\{r\},\{t\},\{s,r\},\{s,t\},\{r,t\},\{s,r,t\}\}$
  yields eight states; most are unreachable from $\{s\}$.  For each
  subset $Q$, $\delta^d(Q,\pmb{a})=\bigcup_{q\in Q}\delta(q,\pmb{a})$ and
  similarly for $\pmb{b}$.  The complete table has 8 rows.

\end{enumerate}

%% 4.4
\item Consider the NDFA from Example~4.2, which has multiple transitions on
$\pmb{0}$: from $s_0$ to both $s_1$ and $s_2$.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (with \textless SOS\textgreater{} and
    \textless EOS\textgreater).}  Encode states with two bits.  The
    nondeterministic transitions become OR gates feeding multiple
    flip-flop inputs simultaneously.  SOS and EOS are incorporated per
    the standard convention.

  \item \emph{DFA via subset construction.}  Starting from the set of
  start states $S_0$, apply $\delta^d$ iteratively.  Since the NDFA
  from Example~4.2 accepts $L=\{x\in\{\pmb{0},\pmb{1}\}^*\mid
  |x|\equiv 0\pmod{2}\vee|x|_{\pmb{1}}\equiv 0\pmod{3}\}$, the
  resulting DFA can be constructed as a product of a two-state machine
  (tracking parity of length) and a three-state machine (tracking
  $|x|_{\pmb{1}}\bmod 3$), giving at most 6 reachable states.
\end{enumerate}

%% 4.5
\item Consider the NDFA from Example~4.3, which accepts the same language as
Example~4.2 but uses multiple start states instead of multiple transitions.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no \textless SOS\textgreater{} or
    \textless EOS\textgreater).}  Each start state is handled by
    OR-ing their outputs into the initial-state signal.

  \item \emph{Circuit diagram (with \textless SOS\textgreater{} and
    \textless EOS\textgreater).}  SOS activates all start states
    simultaneously.

  \item \emph{DFA (connected portion only).}  The start state of $A^d$
  is the set of all start states of $A$.  Applying the subset
  construction yields the connected portion of the DFA.

  \item \emph{Is this DFA isomorphic to any of those in Exercise~4.4?}
  Yes.  Both Example~4.2 and Example~4.3 accept the same language $L$.
  The minimal DFA for $L$ is unique up to isomorphism (by
  Theorem~3.8).  After minimizing the DFA obtained in part~(c) and the
  DFA obtained in Exercise~4.4(b), both must produce the same minimal
  machine (up to renaming of states).  Hence they are isomorphic.
\end{enumerate}

%% 4.6
\item Consider the $\lambda$-NDFA from Example~4.14, with states
$\{s_0,s_1,s_2,s_3\}$ and $\lambda$-transitions $s_0\to s_1$ and $s_1\to s_2$.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram for the original $\lambda$-NDFA (no SOS/EOS).}
  The $\lambda$-transitions are implemented by OR-ing the output of
  the flip-flop encoding $s_i$ directly into the $D$ input of the
  flip-flop encoding $s_j$ whenever $\delta(s_i,\lambda)=\{s_j\}$.
  This creates a combinational path that propagates ``active'' status
  among $\lambda$-connected states within a single clock cycle.

  \item \emph{Circuit diagram for the NDFA from Definition~4.9 (part~(b) of
  Example~4.14, no SOS/EOS).}  After eliminating $\lambda$-transitions using
  $\delta'(s,\pmb{a})=\DBAR_\lambda(s,\pmb{a})$, the resulting NDFA has only
  ordinary transitions; draw the standard NDFA circuit for this machine.
\end{enumerate}

%% 4.7
\item Consider the second NDFA of Example~4.16, which accepts the language
$L^+=\{x_1 x_2\cdots x_k\mid k\ge 1,\, x_i\in L\}$ where $L$ is the language
of strings over $\{\pmb{a},\pmb{b},\pmb{c}\}$ containing exactly one $\pmb{b}$
immediately followed by $\pmb{c}$.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram for $A^{\sigma}_\lambda$ (with EOS but no SOS).}
  Incorporate the $\lambda$-transitions via the standard OR-gate technique
  and add an EOS output gate.

  \item \emph{NDFA without $\lambda$-transitions (Definition~4.9).}
  Compute $\Lambda(s)$ for each state; then set
  $\delta'(s,\pmb{a})=\DBAR_\lambda(s,\pmb{a})$ for all $\pmb{a}\in\Sigma$
  and $s\in S$.  States whose $\lambda$-closure intersects $F$ become final.

  \item \emph{Circuit diagram for the $\lambda$-free NDFA (with EOS, no SOS).}
  Apply the standard circuit encoding to the machine from part~(b).

  \item \emph{DFA (connected portion only).}
  Apply the subset construction to the NDFA from part~(b), starting from
  its set of initial states, and retain only the reachable portion.
\end{enumerate}

%% 4.8
\item \textbf{Complementing the final states of an NDFA does not yield a machine
accepting the complement language.}

For a DFA, $L(\overline{A})=\Sigma^*\setminus L(A)$ because every word follows
exactly one path and is accepted iff it ends in a final state.  Flipping final
and nonfinal states reverses acceptance on that unique path.

For an NDFA, a word $w$ may follow \emph{multiple} paths, some ending in final
states and some not.  Under Definition~4.4, $w\in L(A)$ iff \emph{at least one}
path ends in a final state.  If we complement the final states, a word is
accepted by the modified machine iff at least one path ends in a \emph{formerly
nonfinal} state.  This is \emph{not} the same as $w\notin L(A)$, because a word
can simultaneously have a path to a (formerly) final state and a path to a
(formerly) nonfinal state; the complemented machine would accept it, but it was
already in $L(A)$.

\textbf{Example.}  Let $\Sigma=\{\pmb{a}\}$, $S=\{s_0,s_1\}$, $S_0=\{s_0\}$,
$F=\{s_1\}$, $\delta(s_0,\pmb{a})=\{s_0,s_1\}$, $\delta(s_1,\pmb{a})=\emptyset$.
Then $\pmb{a}\in L(A)$ because $s_1\in\delta(s_0,\pmb{a})$.  After complementing,
$F'=\{s_0\}$, and $\pmb{a}$ is still accepted (via the path $s_0\to s_0\in F'$),
yet $\pmb{a}\notin\Sigma^*\setminus L(A)$ since $\pmb{a}\in L(A)$.  Hence the
complement-of-final-states trick fails for NDFAs.

%% 4.9
\item \textbf{Given an NDFA $A$ without $\lambda$-transitions, construct a
$\lambda$-NDFA $A'$ with exactly one start state, exactly one final state, and
$L(A')=L(A)$.}

\textbf{Construction.}  Let $A=\langle\Sigma,S,S_0,\delta,F\rangle$.  Define
$A'=\langle\Sigma,S',q_0,\delta',\{q_f\}\rangle$ where:
\begin{itemize}
  \item $S'=S\cup\{q_0,q_f\}$ (two new states).
  \item $q_0$ is the unique start state.
  \item $\delta'(q_0,\lambda)=S_0$ (a $\lambda$-transition from $q_0$ to every
        start state of $A$).
  \item For all $s\in S$ and $\pmb{a}\in\Sigma$: $\delta'(s,\pmb{a})=\delta(s,\pmb{a})$.
  \item $\delta'(f,\lambda)\ni q_f$ for every $f\in F$ ($\lambda$-transitions
        from each old final state to the new unique final state $q_f$).
  \item $\delta'(q_f,\pmb{a})=\emptyset$ for all $\pmb{a}$ (and no transitions
        out of $q_f$).
\end{itemize}

\textbf{Correctness.}  A word $w\in\Sigma^+$ is accepted by $A$ iff there is a
path $q_0\xrightarrow{\lambda}s_i\xrightarrow{w}f$ for some $s_i\in S_0$ and
$f\in F$; by the $\lambda$-transition $f\xrightarrow{\lambda}q_f$, this same
path exists in $A'$ ending at $q_f$.  Conversely, any accepting path in $A'$
uses the initial $\lambda$-step to some $s_i\in S_0$, processes $w$ within $S$
(by identical transitions), then uses a final $\lambda$-step to $q_f$, which
corresponds to an accepting path in $A$.  Hence $L(A')=L(A)$.

If $\lambda\in L(A)$ (i.e., some $s\in S_0\cap F$), then we additionally make
$q_0$ a final state.

%% 4.10
\item \textbf{Can part (ii) of Definition~4.8 be deduced from parts (i) and (iii)?}

Definition~4.8 defines $\DBAR$ for NDFAs:
\begin{enumerate}[(i)]
  \item $\DBAR(s,\lambda)=\{s\}$,
  \item $(\forall\pmb{a}\in\Sigma)\,\DBAR(s,\pmb{a})=\delta(s,\pmb{a})$,
  \item $(\forall x\in\Sigma^*)(\forall\pmb{a}\in\Sigma)\,\DBAR(s,x\pmb{a})
        =\bigcup_{t\in\DBAR(s,x)}\delta(t,\pmb{a})$.
\end{enumerate}
Part (ii) concerns single-letter strings.  Setting $x=\lambda$ in part (iii):
\[
  \DBAR(s,\lambda\pmb{a})
  =\bigcup_{t\in\DBAR(s,\lambda)}\delta(t,\pmb{a})
  \;=\;\bigcup_{t\in\{s\}}\delta(t,\pmb{a})
  \;=\;\delta(s,\pmb{a}),
\]
where we used part (i) to obtain $\DBAR(s,\lambda)=\{s\}$ and the fact that
$\lambda\pmb{a}=\pmb{a}$.  Since $\lambda\pmb{a}=\pmb{a}$, this gives
$\DBAR(s,\pmb{a})=\delta(s,\pmb{a})$, which is exactly part (ii).

\textbf{Conclusion.}  Yes, part (ii) can be derived from parts (i) and (iii).
It is included explicitly to make the base case of inductive proofs more
transparent, but it is logically redundant.

%% 4.11
\item \textbf{Alternative construction for removing $\lambda$-transitions.}

Define $S'=S$, $S'_0=\Lambda(S_0)$, $F'=F$, and
$\delta'(s,\pmb{a})=\DBAR_{\lambda}(s,\pmb{a})$ for all $s\in S$, $\pmb{a}\in\Sigma$.

\textbf{Does this work?}  Not in general, because the initial states are wrong.
Setting $S'_0=\Lambda(S_0)$ means the machine starts in the $\lambda$-closure
of the original start states, which is correct for non-empty words.  However, the
problem is with $F'=F$: in the original $\lambda$-NDFA, a state that can reach a
final state via $\lambda$-transitions should \emph{also} be considered final in
the $\lambda$-free version, since accepting the empty string through those
$\lambda$-paths must be respected.  The correct final states should be all
$s\in S$ such that $\Lambda(s)\cap F\neq\emptyset$.

Specifically, if $s_0\xrightarrow{\lambda}\cdots\xrightarrow{\lambda}f\in F$ in
the original machine, then $\lambda\in L(A_\lambda)$.  With $F'=F$, the
$\lambda$-free machine might not accept $\lambda$ (or might accept incorrectly).

\textbf{Fix:}  Replace $F'=F$ with $F'=\{s\in S\mid\Lambda(s)\cap F\neq\emptyset\}$.
With this correction, the construction is equivalent to Definition~4.9 restricted
to $S'=S$ (omitting the extra state $q_0$).

%% 4.12
\item \textbf{Algorithm for union of two FAD languages using $\lambda$-NDFAs.}

Let $L_1=L(A_1)$ and $L_2=L(A_2)$ where $A_1$ and $A_2$ are $\lambda$-NDFAs.
By Exercise~4.9 we may assume each $A_i$ has exactly one start state $q^i_0$ and
exactly one final state $q^i_f$.  Construct $A'$ as follows:
\begin{enumerate}[(1)]
  \item Introduce a new start state $q_0$ and a new final state $q_f$.
  \item Add $\lambda$-transitions: $q_0\xrightarrow{\lambda}q^1_0$ and
        $q_0\xrightarrow{\lambda}q^2_0$.
  \item Add $\lambda$-transitions: $q^1_f\xrightarrow{\lambda}q_f$ and
        $q^2_f\xrightarrow{\lambda}q_f$.
  \item Keep all existing transitions of $A_1$ and $A_2$.
\end{enumerate}
Then $L(A')=L(A_1)\cup L(A_2)$: a word is accepted iff it is accepted by $A_1$
or by $A_2$, because the $\lambda$-transitions route the computation into one
of the two machines and collect the result at $q_f$.

%% 4.13
\item \textbf{Example of NDFA $A$ such that $A^d$ is not connected, not reduced,
but is minimal.}

\textbf{(a) $A^d$ is not connected.}  Any NDFA in which some subsets of states
are never reachable from the start set under the transition function will
produce an $A^d$ with unreachable states.  For example, let
$\Sigma=\{\pmb{a}\}$, $S=\{s_0,s_1\}$, $S_0=\{s_0\}$, $F=\{s_1\}$,
$\delta(s_0,\pmb{a})=\{s_1\}$, $\delta(s_1,\pmb{a})=\emptyset$.
Then $A^d$ has states $\{\emptyset,\{s_0\},\{s_1\},\{s_0,s_1\}\}$ but
$\{s_0,s_1\}$ and $\{s_1\}$ with $\delta^d(\{s_1\},\pmb{a})=\emptyset$ are
reachable, while $\{s_0,s_1\}$ is not reached from $\{s_0\}$.

\textbf{(b) $A^d$ is not reduced.}  $A^d$ contains states (such as
$\emptyset$, the dead state) that are nonfinal and from which no final state is
reachable.  Such states are reachable but do not lead to acceptance; removing
them gives the reduced (trim) machine.

\textbf{(c) $A^d$ is minimal.}  After keeping only the connected, reduced
(trim) portion of $A^d$, the resulting DFA may already be minimal if all
remaining states are distinguishable.  For the example above: the connected,
reduced $A^d$ has states $\{s_0\},\{s_1\},\emptyset$; the states $\{s_0\}$ and
$\emptyset$ are both nonfinal but distinguishable ($\pmb{a}$ from $\{s_0\}$
reaches a final state; from $\emptyset$ it does not), so the machine is minimal.

%% 4.14
\item \textbf{Why was the extra state $q_0$ included in $A^{\sigma}_\lambda$?}

The extra state $q_0$ is needed to handle the special case $\lambda\in L(A_\lambda)$.

In Definition~4.9, the machine $A^{\sigma}_\lambda$ is constructed so that
its start state $q_0$ is \emph{disconnected} from the rest of the machine
(it has no transitions to or from the other states via ordinary letters).
If $\lambda\in L(A_\lambda)$---that is, if the $\lambda$-closure of some
start state of $A_\lambda$ intersects $F$---then $q_0$ is made a final state,
so $\lambda$ is accepted.

Without $q_0$, we would have to designate some ordinary state as the start
state, but any such state already participates in transitions that process
non-empty strings.  Making it simultaneously responsible for the acceptance of
$\lambda$ can create conflicts.

\textbf{Example.}  Suppose $A_\lambda$ has start state $s_0$ with
$s_0\in F$ (so $\lambda\in L(A_\lambda)$) and transitions that extend the
language.  If we set $q_0=s_0$ without creating a separate new start state,
the $\lambda$-closure would ``spread'' acceptance to additional states in ways
that might incorrectly mark non-$\lambda$ prefixes as accepted.  The fresh
isolated $q_0$ cleanly separates the ``zero-length acceptance'' case from the
rest of the computation.

%% 4.15
\item \textbf{Union construction using NDFAs \emph{without} $\lambda$-transitions.}

\begin{enumerate}[(a)]
  \item \textbf{Algorithm.}  Let $A_1=\langle\Sigma,S_1,S^1_0,\delta_1,F_1\rangle$
  and $A_2=\langle\Sigma,S_2,S^2_0,\delta_2,F_2\rangle$ (assume $S_1\cap S_2=\emptyset$).
  Construct $A'=\langle\Sigma,S_1\cup S_2,S^1_0\cup S^2_0,\delta',F_1\cup F_2\rangle$
  where $\delta'(s,\pmb{a})=\delta_i(s,\pmb{a})$ for $s\in S_i$.
  The machine $A'$ starts in the combined set of initial states and accepts
  any word accepted by either $A_1$ or $A_2$.  Thus $L(A')=L(A_1)\cup L(A_2)$.

  \item \textbf{Comparison.}  This is actually \emph{easier} than the
  $\lambda$-transition approach (Exercise~4.12): no new states are needed at
  all, since we simply merge the state sets and initial-state sets.

  \item \textbf{Can we ensure exactly one start state and one final state?}
  Exactly one start state: \emph{not in general} without adding a new
  state or using $\lambda$-transitions.  If we introduce a new state $q_0$
  and require $q_0$ to mimic all transitions from every $s\in S^1_0\cup S^2_0$
  (setting $\delta'(q_0,\pmb{a})=\delta'(s,\pmb{a})$ for each $s$ in the
  original start sets), we can achieve a single start state---but we must also
  make $q_0$ final if any original start state was final (to handle $\lambda$).
  Exactly one final state can similarly be achieved by merging: add a new
  state $q_f$ and, for each $f\in F_1\cup F_2$, redirect: whenever $f$ would
  have been final, instead make $q_f$ absorb $f$'s role by duplicating its
  transitions.  Without $\lambda$-transitions this is more complex than
  the $\lambda$-based construction.
\end{enumerate}

%% 4.16
\item Consider the $\lambda$-NDFA $A^{\sigma}_\lambda$ from Example~4.15 (the
machine without $\lambda$-transitions corresponding to the automaton of
Example~4.14).

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no SOS or EOS).}
  From Example~4.14, the $\lambda$-closures are
  $\Lambda(s_0)=\{s_0,s_1,s_2\}$, $\Lambda(s_1)=\{s_1,s_2\}$,
  $\Lambda(s_2)=\{s_2\}$, $\Lambda(s_3)=\{s_3\}$.
  The transition function of $A^{\sigma}_\lambda$ (without $\lambda$-moves)
  is $\delta^{\sigma}_\lambda(s,\pmb{a})=\DBAR_\lambda(s,\pmb{a})$.
  Encode states with two bits; draw next-state logic using OR gates for
  nondeterministic branches, following the method of Chapter~4.

  \item \emph{Convert to $A^{\sigma d}_\lambda$ (connected portion only).}
  Apply the subset construction to $A^{\sigma}_\lambda$.  Starting from the
  initial set determined by $S_0=\{q_0\}$ (or the appropriate start state
  of $A^{\sigma}_\lambda$), compute $\delta^d$ for each reachable subset and
  each input letter, retaining only connected states.
\end{enumerate}

%% 4.17
\item \begin{enumerate}[(a)]
  \item \textbf{Proof that $\DBAR(s,\pmb{a})=\delta(s,\pmb{a})$ for any NDFA
  without $\lambda$-transitions.}

  By Definition~4.8(ii), $\DBAR(s,\pmb{a})=\delta(s,\pmb{a})$ for all
  $s\in S$ and $\pmb{a}\in\Sigma$.  This is immediate from the definition;
  alternatively, apply part (iii) with $x=\lambda$:
  \[
    \DBAR(s,\lambda\pmb{a})
    =\bigcup_{t\in\DBAR(s,\lambda)}\delta(t,\pmb{a})
    =\bigcup_{t\in\{s\}}\delta(t,\pmb{a})
    =\delta(s,\pmb{a}),
  \]
  using part (i): $\DBAR(s,\lambda)=\{s\}$.  Since $\lambda\pmb{a}=\pmb{a}$,
  we conclude $\DBAR(s,\pmb{a})=\delta(s,\pmb{a})$. \qed

  \item \textbf{Counterexample for $\lambda$-NDFAs.}

  Let $\Sigma=\{\pmb{a}\}$, $S=\{s_0,s_1\}$, $S_0=\{s_0\}$, $F=\{s_1\}$,
  $\delta_\lambda(s_0,\lambda)=\{s_1\}$, $\delta_\lambda(s_0,\pmb{a})=\emptyset$,
  $\delta_\lambda(s_1,\pmb{a})=\emptyset$.

  Then $\Lambda(s_0)=\{s_0,s_1\}$.  So
  \[
    \DBAR_\lambda(s_0,\pmb{a})
    =\Lambda\!\Bigl(\bigcup_{t\in\Lambda(s_0)}\delta_\lambda(t,\pmb{a})\Bigr)
    =\Lambda(\emptyset\cup\emptyset)=\emptyset,
  \]
  but $\delta_\lambda(s_0,\pmb{a})=\emptyset$ also.  Let us try a different
  example: $\delta_\lambda(s_0,\lambda)=\{s_1\}$, $\delta_\lambda(s_1,\pmb{a})=\{s_1\}$.
  Then $\delta_\lambda(s_0,\pmb{a})=\emptyset$, yet
  \[
    \DBAR_\lambda(s_0,\pmb{a})
    =\Lambda\!\Bigl(\bigcup_{t\in\{s_0,s_1\}}\delta_\lambda(t,\pmb{a})\Bigr)
    =\Lambda(\{s_1\})=\{s_1\}\neq\emptyset=\delta_\lambda(s_0,\pmb{a}).
  \]
  This demonstrates that $\DBAR_\lambda(s,\pmb{a})\neq\delta_\lambda(s,\pmb{a})$
  is possible for $\lambda$-NDFAs.
\end{enumerate}

%% 4.18
\item Consider the NDFA accepting the original language $L$ from Example~4.10:
$L=\{x\in\{\pmb{a},\pmb{b}\}^*\mid x\text{ begins with }\pmb{a}\vee x\text{ contains }\pmb{ba}\}$.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no SOS/EOS).}  The NDFA has a start state
  with multiple transitions and states tracking whether the substring $\pmb{ba}$
  has been seen.  Encode states with the appropriate number of bits and draw
  the next-state logic using OR gates for nondeterministic branching.

  \item \emph{DFA (connected portion only).}  Apply the subset construction
  to the NDFA, starting from the set of initial states.  The language $L$ is
  a union of two simpler languages ($x$ begins with $\pmb{a}$, and $x$ contains
  $\pmb{ba}$), each recognized by small DFAs; the resulting product DFA will
  have at most $2\times 3=6$ states (two states for ``starts with $\pmb{a}$''
  and three for tracking the substring $\pmb{ba}$).
\end{enumerate}

%% 4.19
\item Consider the NDFA accepting $L^r=\{x\in\{\pmb{a},\pmb{b}\}^*\mid x\text{ ends with }\pmb{a}\vee x\text{ contains }\pmb{ab}\}$ from Example~4.10.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no SOS/EOS).}  This NDFA is obtained by
  reversing all arrows and swapping start and final states of the NDFA for $L$.
  The multiple start states of $L^r$'s NDFA become multiple flip-flop inputs
  activated at initialization.

  \item \emph{DFA (connected portion only).}  Apply the subset construction.
  Because $L^r$ is the reverse of $L$ (which is FAD), $L^r$ is also FAD by
  Theorem~5.9 (closure under reversal).  The minimal DFA for $L^r$ can be
  computed from the subset construction applied to the reversed NDFA.
\end{enumerate}

%% 4.20
\item \textbf{Are the pumping lemma arguments still valid for NDFAs?}

Yes.  The pumping lemma (Chapter~2) applies to any language accepted by a DFA
(or equivalently, any FAD language).  Since every NDFA $A$ is equivalent to a
DFA $A^d$ (Theorem~4.1), $L(A)=L(A^d)$ is FAD, and the pumping lemma applies
to $L(A)$.

More directly: the pumping lemma is a property of the \emph{language}, not the
machine representation.  If $L$ is FAD (accepted by some NDFA), then all the
arguments from Chapter~2 go through unchanged.  In particular, if $A$ has $n$
states, then $A^d$ has at most $2^n$ states; words of length at least $2^n$
must cause a repeated state in $A^d$, yielding the pumping decomposition.

%% 4.21
\item \textbf{Does the conclusion of Theorem~2.7 still hold for NDFAs?}

Yes.  Theorem~2.7 characterizes FAD languages via the Myhill--Nerode theorem:
$L$ is FAD iff $\equiv_L$ has finitely many equivalence classes.  This is a
property of the language $L$, not of the particular automaton (DFA or NDFA)
accepting it.  Since NDFAs and DFAs accept exactly the same class of languages
(the FAD languages, by Theorem~4.1), the conclusion holds: a language is
accepted by some NDFA iff it is accepted by some DFA iff $\equiv_L$ has
finitely many classes iff the minimal DFA is finite.

%% 4.22
\item \textbf{Given an NDFA $A$ without $\lambda$-transitions with $\lambda\notin L(A)$,
construct $A''$ with exactly one start state, one final state, and $L(A'')=L(A)$.}

\textbf{Construction.}  Let $A=\langle\Sigma,S,S_0,\delta,F\rangle$ with
$\lambda\notin L(A)$ (i.e., $S_0\cap F=\emptyset$).
\begin{enumerate}[1.]
  \item Add a new start state $q_0$ and a new final state $q_f$.
  \item For all $\pmb{a}\in\Sigma$:
        $\delta''(q_0,\pmb{a})=\bigcup_{s\in S_0}\delta(s,\pmb{a})$.
        (Simulate starting from any state in $S_0$.)
  \item For all $s\in S$ and $\pmb{a}\in\Sigma$:
        $\delta''(s,\pmb{a})=\delta(s,\pmb{a})$.
  \item For all $f\in F$: whenever $f$ would be in $\delta''(s,\pmb{a})$
        for some $s,\pmb{a}$, also add $q_f$ to $\delta''(s,\pmb{a})$.
        (That is, $q_f$ behaves like a ``copy'' of states in $F$:
        $\delta''(s,\pmb{a})=(\delta(s,\pmb{a})\setminus F)\cup\{q_f\mid\delta(s,\pmb{a})\cap F\neq\emptyset\}$.)
  \item $\delta''(q_f,\pmb{a})=\emptyset$ for all $\pmb{a}$.
\end{enumerate}
The new machine $A''$ has start state $q_0$, final state $\{q_f\}$, and
accepts a word $w=\pmb{a}_1\cdots\pmb{a}_k$ ($k\ge 1$) iff $A$ accepts $w$.
Since $\lambda\notin L(A)$, we need $S_0\cap F=\emptyset$, ensuring $q_0$ need
not be final.

%% 4.23
\item \textbf{If $\lambda\in L(A)$, the three conditions of Exercise~4.22
may be incompatible (without $\lambda$-transitions).}

\textbf{Example.}  Let $\Sigma=\{\pmb{a}\}$, $S=\{s_0\}$, $S_0=\{s_0\}$,
$F=\{s_0\}$, $\delta(s_0,\pmb{a})=\{s_0\}$.  Then $L(A)=\{\pmb{a}\}^*$, which
includes $\lambda$.

We want $A''$ with: (1) one start state $q_0$; (2) one final state $q_f$; (3)
$L(A'')=\{\pmb{a}\}^*$.

For $\lambda\in L(A'')$, we need $q_0\in F''$, i.e., $q_0=q_f$.  But if
$q_0=q_f$ is the unique start-and-final state, then for $A''$ to accept $\pmb{a}$,
we need a transition from $q_f$ to some state that eventually reaches $q_f$---
but $q_f$ is the only state, so $\delta''(q_f,\pmb{a})=\{q_f\}$.  This works
for this particular example.

The general obstruction arises when $L(A)$ requires the single final state to
also be a non-trivial \emph{source} of transitions, while the single start state
needs to be the same final state.  In machines where $\lambda\in L(A)$ but
words of positive length follow different paths from the start state to
the final state, merging start and final forces all transitions through a single
state that must simultaneously act as the ``beginning'' and ``end'', which may
require $\lambda$-transitions to route correctly.  Without $\lambda$-transitions,
it may be impossible to separate the acceptance of $\lambda$ (at the start
state) from the acceptance of longer words (which require reaching $q_f$ after
processing letters).

%% 4.24
\item \textbf{Prove Lemma~4.2.}

\textbf{Lemma~4.2:} For any $s\in S$ and any $x\in\Sigma^+$,
$\overline{\delta^{\sigma}_\lambda}(s,x)=\DBAR_\lambda(s,x)$.

\begin{proof}
We proceed by induction on $|x|=k\ge 1$.

\textbf{Base case ($k=1$).}  Let $x=\pmb{a}\in\Sigma$.  By definition of
$\delta^{\sigma}_\lambda$:
\[
  \delta^{\sigma}_\lambda(s,\pmb{a})
  =\Lambda\!\Bigl(\bigcup_{t\in\Lambda(s)}\delta_\lambda(t,\pmb{a})\Bigr)
  =\DBAR_\lambda(s,\pmb{a}).
\]
So $\overline{\delta^{\sigma}_\lambda}(s,\pmb{a})=\delta^{\sigma}_\lambda(s,\pmb{a})=\DBAR_\lambda(s,\pmb{a})$.

\textbf{Inductive step.}  Assume the result holds for all words of length $k$.
Let $y=x\pmb{a}$ with $|x|=k$ and $\pmb{a}\in\Sigma$.  Then:
\begin{align*}
  \overline{\delta^{\sigma}_\lambda}(s,x\pmb{a})
  &= \bigcup_{t\in\overline{\delta^{\sigma}_\lambda}(s,x)}
      \delta^{\sigma}_\lambda(t,\pmb{a}) \\
  &= \bigcup_{t\in\DBAR_\lambda(s,x)}
      \delta^{\sigma}_\lambda(t,\pmb{a}) \quad\text{(induction hypothesis)} \\
  &= \bigcup_{t\in\DBAR_\lambda(s,x)}
      \DBAR_\lambda(t,\pmb{a}) \quad\text{(base case)} \\
  &= \DBAR_\lambda(s,x\pmb{a}).
    \quad\text{(definition of $\DBAR_\lambda$)}
\end{align*}
By induction, the result holds for all $x\in\Sigma^+$. \qed
\end{proof}

%% 4.25
\item \textbf{A DFA can be viewed as an NDFA, and $L(A^n)=L(A)$.}

\textbf{Construction.}  Given a DFA $A=\langle\Sigma,S,s_0,\delta,F\rangle$
(with $\delta:S\times\Sigma\to S$ a total function), define the NDFA
$A^n=\langle\Sigma,S,\{s_0\},\delta^n,F\rangle$ where
$\delta^n:S\times\Sigma\to\wp(S)$ is given by $\delta^n(s,\pmb{a})=\{\delta(s,\pmb{a})\}$.

$A^n$ is a valid NDFA: $S$ is finite, $S_0^n=\{s_0\}$ is a nonempty set of
start states, and $\delta^n$ maps to subsets of $S$.

\textbf{Claim: $L(A^n)=L(A)$.}

\begin{proof}
By induction on $|x|$, we show $\DBAR^n(s_0,x)=\{\deltabar{}(s_0,x)\}$ (i.e., the
nondeterministic extended transition function applied to the singleton start
always yields the singleton $\{\deltabar{}(s_0,x)\}$).

\emph{Base:} $\DBAR^n(s_0,\lambda)=\{s_0\}=\{\deltabar{}(s_0,\lambda)\}$.

\emph{Step:} $\DBAR^n(s_0,x\pmb{a})
=\bigcup_{t\in\DBAR^n(s_0,x)}\delta^n(t,\pmb{a})
=\bigcup_{t\in\{\deltabar{}(s_0,x)\}}\{\delta(t,\pmb{a})\}
=\{\delta(\deltabar{}(s_0,x),\pmb{a})\}
=\{\deltabar{}(s_0,x\pmb{a})\}$.

Hence $x\in L(A^n)$ iff $\DBAR^n(s_0,x)\cap F\neq\emptyset$ iff
$\deltabar{}(s_0,x)\in F$ iff $x\in L(A)$. \qed
\end{proof}

%% 4.26
\item \textbf{Show that $\overline{\delta^{\sigma}_\lambda}(s,\lambda)\neq\DBAR_\lambda(s,\lambda)$ in general.}

By definition:
\begin{itemize}
  \item $\overline{\delta^{\sigma}_\lambda}(s,\lambda)=\{s\}$ (definition of
    the extended transition function for zero-length input).
  \item $\DBAR_\lambda(s,\lambda)=\Lambda(s)$ (definition of $\DBAR_\lambda$ for $\lambda$).
\end{itemize}

These differ whenever $\Lambda(s)\neq\{s\}$, i.e., whenever there exists at
least one $\lambda$-transition out of $s$.

\textbf{Example.}  In Example~4.14, $\Lambda(s_0)=\{s_0,s_1,s_2\}$ but
$\overline{\delta^{\sigma}_\lambda}(s_0,\lambda)=\{s_0\}$.  Hence
$\overline{\delta^{\sigma}_\lambda}(s_0,\lambda)\neq\DBAR_\lambda(s_0,\lambda)$.

This is why Lemma~4.2 is stated only for $x\in\Sigma^+$ (words of length
$\ge 1$) and cannot be extended to $\lambda$.

%% 4.27
\item \textbf{Prove: $(\forall A,B\in\wp(S))(\forall\pmb{a}\in\Sigma)\,
\delta^d(A\cup B,\pmb{a})=\delta^d(A,\pmb{a})\cup\delta^d(B,\pmb{a})$.}

\begin{proof}
By Definition~4.5, $\delta^d(Q,\pmb{a})=\bigcup_{q\in Q}\delta(q,\pmb{a})$
for any $Q\in\wp(S)$.  Therefore:
\begin{align*}
  \delta^d(A\cup B,\pmb{a})
  &= \bigcup_{q\in A\cup B}\delta(q,\pmb{a}) \\
  &= \Bigl(\bigcup_{q\in A}\delta(q,\pmb{a})\Bigr)\cup\Bigl(\bigcup_{q\in B}\delta(q,\pmb{a})\Bigr) \\
  &= \delta^d(A,\pmb{a})\cup\delta^d(B,\pmb{a}). \qed
\end{align*}
\end{proof}

%% 4.28
\item \textbf{Acceptance requiring \emph{every} reachable state to be final.}

\begin{enumerate}[(a)]
  \item \textbf{New acceptance definition.}  Define
  $L_{\forall}(A)=\{w\in\Sigma^*\mid\DBAR(S_0,w)\neq\emptyset\wedge\DBAR(S_0,w)\subseteq F\}$.
  That is, $w$ is accepted iff (i) at least one state is reachable and
  (ii) \emph{every} reachable state is final.

  \item \textbf{Modified $A^d$.}  Define $A^{\forall}=A^d$ with the same
  state set $\wp(S)$ and transition function $\delta^d$ as before, but with
  the new final-state condition:
  \[
    F^{\forall}=\{Q\in\wp(S)\mid Q\neq\emptyset\wedge Q\subseteq F\}.
  \]
  That is, $Q$ is a final state of $A^{\forall}$ iff $Q$ is nonempty and
  all states in $Q$ are final states of $A$.  By the same argument as
  Theorem~4.1 (modified for this new acceptance criterion),
  $L_{\forall}(A)=L(A^{\forall})$.

  Note: this ``universal'' acceptance is still FAD (the modified $A^{\forall}$
  is a DFA), so $L_{\forall}(A)$ is always a FAD language.
\end{enumerate}

%% 4.29
\item \textbf{Draw the connected part of $T^d$, the deterministic equivalent of
the NDFA $T$ in Example~4.7.}

The NDFA $T$ from Example~4.7 monitors a bit stream $\{\pmb{0},\pmb{1}\}^*$ and
enters a final state when the substring $\pmb{010010}$ is detected.  $T$ has a
``backbone'' chain of states tracking progress toward matching $\pmb{010010}$,
plus loops back on mismatch.

To compute $T^d$: starting from $S_0=\{s_0\}$ (the initial set), compute
$\delta^d$ for $\pmb{0}$ and $\pmb{1}$.  The key observation is that the NDFA
has the initial state $s_0$ loop on all inputs (to allow $\pmb{010010}$ to occur
as a \emph{substring}), so $s_0$ is always in every reachable state-set of $T^d$.
This means $T^d$ has relatively few distinct reachable state-sets.

The connected portion of $T^d$ can be systematically computed: from each
state-set $Q$, compute $\delta^d(Q,\pmb{0})$ and $\delta^d(Q,\pmb{1})$, where
$\delta^d(Q,\pmb{b})=\bigcup_{q\in Q}\delta(q,\pmb{b})$.  Since $s_0$
self-loops on both inputs and $T$'s non-start states progress or reset toward
the pattern, the resulting DFA has at most $2^7$ states but in practice far fewer
are reachable.

%% 4.30
\item \textbf{Modify $T$ to revert to a nonfinal state when $\pmb{000111}$ (EOT) is detected.}

Extend $T$ with additional states tracking progress toward matching $\pmb{000111}$
from within a final (recording) state.  Once in a final state $f$:
\begin{itemize}
  \item If the incoming bits form $\pmb{000111}$, the machine returns to a
  designated nonfinal ``idle'' state.
  \item Otherwise, it remains in a final state.
\end{itemize}
Formally, add a chain of 6 states $f_1,f_2,f_3,f_4,f_5,f_6$ (all final, since
recording is still active during $\pmb{000111}$), where $f_i$ represents having
matched the first $i$ characters of $\pmb{000111}$ while recording.  If all 6
are matched, transition to a nonfinal state.  On mismatch at any $f_i$, return
to the recording state $f$ (or $f_1$ if the failed character is still the start
of a potential new EOT signal).  The resulting NDFA now accepts the stream
while it is ``recording'' and stops recording upon detecting $\pmb{000111}$.

%% 4.31
\item Consider the NDFA $A$ from Example~4.14.

\begin{enumerate}[(a)]
  \item \textbf{Diagram of $A^{\sigma}_\lambda$.}
  From the text, $A_\lambda$ (Example~4.14) has states $\{s_0,s_1,s_2,s_3\}$
  with $\lambda$-transitions $s_0\to s_1$ and $s_1\to s_2$.  By Definition~4.9,
  $A^{\sigma}_\lambda$ has the same state set plus an extra start state $q_0$
  (disconnected unless $\lambda\in L$), and transitions
  $\delta^{\sigma}_\lambda(s,\pmb{a})=\DBAR_\lambda(s,\pmb{a})$ for each
  $s\in\{s_0,s_1,s_2,s_3\}$ and $\pmb{a}\in\Sigma$.

  For example, if the original machine has $\delta_\lambda(s_2,\pmb{a})=\{s_3\}$
  and $\delta_\lambda(s_0,\pmb{a})=\emptyset$, then:
  \[
    \delta^{\sigma}_\lambda(s_0,\pmb{a})
    =\Lambda\!\Bigl(\bigcup_{t\in\{s_0,s_1,s_2\}}\delta_\lambda(t,\pmb{a})\Bigr)
    =\Lambda(\{s_3\})=\{s_3\}.
  \]

  \item \textbf{Diagram of $A^{\sigma d}_\lambda$ (connected portion).}
  Apply the subset construction to $A^{\sigma}_\lambda$, starting from
  $\{q_0\}$ (or the appropriate initial state set), and draw only the reachable
  state-sets.
\end{enumerate}

%% 4.32
\item \textbf{What is wrong with the given ``proof'' of Lemma~4.2?}

The flaw is in the inductive step.  The ``proof'' starts the induction at $k=1$
(base case for single letters), but the inductive step uses
$\overline{\delta^{\sigma}_\lambda}(s,x)$ for $x\in\Sigma^k$ and then applies
the ``induction hypothesis'' $\overline{\delta^{\sigma}_\lambda}(s,x)=\DBAR_\lambda(s,x)$.

The critical error is on the line:
\[
  \delta^{\sigma}_\lambda(\overline{\delta^{\sigma}_\lambda}(s,x),\pmb{a})
  = \overline{\delta^{\sigma}_\lambda}(\DBAR_\lambda(s,x),\pmb{a}).
\]
Here, $\overline{\delta^{\sigma}_\lambda}(s,x)$ is a \emph{set} of states
(since we're in an NDFA), but $\delta^{\sigma}_\lambda$ is defined on individual
states, not sets.  The expression $\delta^{\sigma}_\lambda(Q,\pmb{a})$ where
$Q$ is a set is not directly the same as $\overline{\delta^{\sigma}_\lambda}(Q,\pmb{a})$
for a set-valued argument: the extended function $\overline{\delta^{\sigma}_\lambda}$
on a set-argument must take the union over all states in the set.

More specifically, the step
\[
  \delta^{\sigma}_\lambda\!\bigl(\overline{\delta^{\sigma}_\lambda}(s,x),\pmb{a}\bigr)
  = \overline{\delta^{\sigma}_\lambda}\!\bigl(\DBAR_\lambda(s,x),\pmb{a}\bigr)
\]
conflates applying $\delta^{\sigma}_\lambda$ to a \emph{set} with applying the
set-union version of $\overline{\delta^{\sigma}_\lambda}$ to a set argument.  These
are not the same: the left side applies $\delta^{\sigma}_\lambda$ to the entire
set at once (which is not how the function is defined), while the right side
correctly takes the union over individual states.  This makes the step invalid
without justification.

%% 4.33
\item Consider the NDFA $T$ from Example~4.7 (detecting substring $\pmb{010010}$).

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no SOS/EOS).}  $T$ has 8 states (one start
  state that loops on all input plus 7 states tracking progress through
  $\pmb{010010}$); encode with 3 bits.  The nondeterminism (the start state
  $s_0$ loops while simultaneously ``starting'' to track the pattern)
  is implemented by OR-ing the self-loop into the initial state's $D$ input.

  \item \emph{DFA $T^d$ (connected portion).}  See Exercise~4.29 for the
  subset construction.  The resulting DFA recognizes the same language as $T$.
\end{enumerate}

%% 4.34
\item Consider the NDFA from Example~4.11, which accepts all words containing
$\pmb{10110}$, $\pmb{1010}$, or $\pmb{01101}$ as substrings.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram (no SOS/EOS).}  The NDFA has a single initial
  state $s_0$ that loops on $\Sigma^*$, and three parallel ``arms''---one for
  each target substring.  Encode all states; the circuit feeds the output of the
  initial state into each arm simultaneously.  Final state output is the OR of
  the final states of all three arms (and the appended $\Sigma^*$-accepting tail
  $S_{18}$).

  \item \emph{DFA (connected portion only).}  Since all three patterns are
  independent, the subset construction produces a DFA whose states track
  progress through each arm simultaneously.  The reachable state-sets are
  bounded and the connected portion is much smaller than $2^{|S|}$.
\end{enumerate}

%% 4.35
\item Consider the NDFA $B$ from Example~4.8 and its deterministic equivalent $B^d$.

\begin{enumerate}[(a)]
  \item \emph{Circuit diagram for $B$ (no SOS/EOS).}  Encode the states of $B$
  using $\lceil\log_2|S_B|\rceil$ bits.  The nondeterministic transitions become
  OR gates feeding multiple $D$ flip-flop inputs simultaneously, as described in
  Chapter~4.

  \item \emph{Circuit diagram for $B^d$ (no SOS/EOS).}  The states of $B^d$ are
  subsets of the states of $B$.  Encode them using as many bits as there are
  states in $B$ (one bit per state of $B$, where each bit indicates membership
  in the current subset).  This ``one-hot-like'' encoding makes the circuit for
  $B^d$ structurally similar to that of $B$: the next-state logic for bit $j$
  (corresponding to state $s_j$ of $B$) in $B^d$ is the OR of the $\delta$
  contributions from all currently active states into $s_j$.  This is exactly
  the same combinational logic as for $B$, confirming the structural similarity.
\end{enumerate}

%% 4.36
\item \textbf{Circuit diagrams for each NDFA in Figure~4.25 (no SOS/EOS).}

For each NDFA $A_i$ in Figure~4.25, with $\Sigma=\{\pmb{a},\pmb{b},\pmb{c}\}$:
\begin{enumerate}[1.]
  \item Assign a bit-vector encoding to each state.
  \item For each state $s_j$ and each input letter $\pmb{\sigma}$, compute the
    set of states $\delta(s_j,\pmb{\sigma})$.
  \item The $D$ input of the flip-flop for state $s_k$ is the OR of the outputs
    of all flip-flops $s_j$ for which $s_k\in\delta(s_j,\pmb{\sigma}_{current})$.
  \item The output is the OR of all final-state flip-flops.
\end{enumerate}
(Specific gate-level diagrams depend on the transitions shown in Figure~4.25.)

%% 4.37
\item \textbf{Circuit diagrams for each NDFA in Figure~4.25 (with SOS and EOS).}

Extend each circuit from Exercise~4.36 with:
\begin{itemize}
  \item \textbf{SOS input:} an asynchronous reset that forces all flip-flops to
    the encoding of the start state(s) simultaneously.
  \item \textbf{EOS input:} gates the final-state output signal, so the
    ``accept'' output is only asserted during the EOS pulse, not continuously.
\end{itemize}
See Example~4.13 for the solution corresponding to Figure~4.25e.

%% 4.38
\item \textbf{Adapting Definition~3.10 for NDFAs and proving that an algorithm
exists to find the connected portion.}

\begin{enumerate}[(a)]
  \item \textbf{Adapted definition.}  The connected portion of an NDFA
  $A=\langle\Sigma,S,S_0,\delta,F\rangle$ is defined as follows.  Let
  \[
    \text{Reach}_0 = S_0,\qquad
    \text{Reach}_{k+1} = \text{Reach}_k
      \cup\bigcup_{\pmb{a}\in\Sigma}\bigcup_{s\in\text{Reach}_k}\delta(s,\pmb{a}).
  \]
  The connected portion is $\text{Reach}=\bigcup_{k\ge 0}\text{Reach}_k$, together
  with all transitions among states in $\text{Reach}$ and the restriction of
  $F$ to $\text{Reach}$.

  \item \textbf{Proof that the algorithm terminates.}

  \begin{proof}
  The sets $\text{Reach}_k$ form a non-decreasing chain
  $\text{Reach}_0\subseteq\text{Reach}_1\subseteq\cdots\subseteq S$.
  Since $S$ is finite, the chain must stabilize: there exists $k^*$ such that
  $\text{Reach}_{k^*}=\text{Reach}_{k^*+1}$.  At that point, no new states are
  added.  The algorithm terminates in at most $|S|$ iterations (since at least
  one new state must be added in each non-stable step, and there are only $|S|$
  states).

  To compute this algorithmically:
  \begin{enumerate}[1.]
    \item Initialize $R:=S_0$, $\text{New}:=S_0$.
    \item While $\text{New}\neq\emptyset$:
      \begin{enumerate}[(a)]
        \item $\text{Next}:=\bigcup_{\pmb{a}\in\Sigma}\bigcup_{s\in\text{New}}
              \delta(s,\pmb{a})\setminus R$.
        \item $R:=R\cup\text{Next}$; $\text{New}:=\text{Next}$.
      \end{enumerate}
    \item Output $R$ as the connected portion.
  \end{enumerate}
  Each iteration of the while loop adds at least one state to $R$ (otherwise
  $\text{New}=\emptyset$ and we stop).  Thus the loop executes at most $|S|$
  times, and the algorithm runs in $O(|S|\cdot|\Sigma|)$ time. \qed
  \end{proof}
\end{enumerate}

\end{enumerate}
